---
title: "Biostat 203B Homework 1"
author: Dominic Shehtanian
subtitle: Due Jan 22 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email.

**Solution:** Done.


2. Create a **private** repository `biostat-203b-2021-winter` and add `Hua-Zhou`, `Chris-German` and `ElvisCuiHan` as your collaborators with write permission.

**Solution:** Done.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `master` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `master` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and data sets to reproduce results) in `master` branch.

**Solution:** Done.

4. After each homework due date, teaching assistant and instructor will check out your master branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

**Solution:** Done.

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

**Solution:** Acknowledged.

## Q2. Linux Shell Commands

1. This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic-iv.mit.edu/docs/access/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

**Solution:** 

[Course Completion](https://www.citiprogram.org/verify/?k2d22ea07-a433-46a8-ad5a-6a92d7fec6b0-40373283)

[Certificate](https://www.citiprogram.org/verify/?wdf218c49-bd83-40e6-89b3-3d4f2f64d125-40373283)

2. The `/usr/203b-data/mimic-iv/` folder on teaching server contains data sets from MIMIC-IV. Refer to <https://mimic-iv.mit.edu/docs/datasets/> for details of data files.  

    ```{bash}
    ls -l /usr/203b-data/mimic-iv/
    ```

Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/usr/203b-data/mimic-iv` directly in following exercises. 

**Solution:** Acknowledged.

    Use Bash commands to answer following questions.

3. Display the contents in the folders `core`, `hosp`, `icu`. What are the functionalities of the bash commands `zcat`, `zless`, `zmore`, and `zgrep`? 

**Solution:**
```{bash}
ls -l /usr/203b-data/mimic-iv/core /usr/203b-data/mimic-iv/hosp \
/usr/203b-data/mimic-iv/icu
```

zcat - Prints the contents of files which are compressed.

zmore - Allows reading through text of compressed files one screen at a time.
        Only allows for forward movement.
        
zless - Allows reading through text of compressed files one screen at a time.
        Allows for forward and backward movement and scrolling.
        
zgrep - Allows for searching compressed files for strings/regular expressions.

4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        ls -l $datafile
      done
    ```
**Solution:** This loop runs the "ls -l" command for each file ending in .gz in the /usr/203b-data/mimic-iv/core/ directory.

Display the number of lines in each data file using a similar loop.

**Solution:** 

    ```{bash}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        wc -l $datafile
      done
    ```

5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, `language`, `marital_status`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq`, `wc`, and so on.)

**Solution:** 
Displaying the first few lines of the file.
```{bash}
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | head
```
Number of rows in the data file and number of unique patients.
```{bash}
  echo "# of rows:" 
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz | wc -l
  echo "# of unique patients:"
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $1}' | tail -n+2 | sort -u | wc -l
```
Possible values taken by admission_type.
```{bash}
  echo "Possible values of" 
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $6}' | head -n 1
  echo ''
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $6}' | tail -n+2 | sort | uniq -c
```
Possible values taken by admission_location.
```{bash}
  echo "Possible values of" 
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $7}' | head -n 1
  echo ''
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $7}' | tail -n+2 | sort | uniq -c
```
Possible values taken by insurance.
```{bash}
  echo "Possible values of" 
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $9}' | head -n 1
  echo ''
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $9}' | tail -n+2 | sort | uniq -c
```
Possible values taken by language.
```{bash}
  echo "Possible values of" 
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $10}' | head -n 1
  echo ''
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $10}' | tail -n+2 | sort | uniq -c
```
Possible values taken by marital_status.
```{bash}
  echo "Possible values of" 
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $11}' | head -n 1
  echo ''
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $11}' | tail -n+2 | sort | uniq -c
```
Possible values taken by ethnicity.
```{bash}
  echo "Possible values of" 
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $12}' | head -n 1
  echo ''
  zcat /usr/203b-data/mimic-iv/core/admissions.csv.gz \
  | awk -F, '{print $12}' | tail -n+2 | sort | uniq -c
```

## Q3. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    curl http://www.gutenberg.org/cache/epub/42671/pg42671.txt > pride_and_prejudice.txt
    ```

Do **not** put this text file `pride_and_prejudice.txt` in Git. Using a `for` loop, how would you tabulate the number of times each of the four characters is mentioned?

**Solution:**
```{bash}
    for name in {Elizabeth,Jane,Lydia,Darcy}
      do
        ment=$(grep -o $name ./pride_and_prejudice.txt | wc -l)
        echo "$name is mentioned $ment times."
      done
```
Elizabeth is mentioned the most.

2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```

**Solution:**
  The first command writes 'hello, world' (without commas, in every case mentioned in this solution) to a file. Each time the command is run, the file is overwritten and only contains one line reading 'hello, world'.
  The second command appends 'hello, world' to a file. Each time the command is run, a new line containing 'hello, world' is added to the end of the file.
  
3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash}
    chmod 770 ./middle.sh
    ./middle.sh pride_and_prejudice.txt 20 5
    ```

Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

**Solution:** This script outputs a section from the middle of a file, in this case lines 16-20 of the text of Pride and Prejudice. It first outputs the first 20 lines of pride_and_prejudice.txt, and then it outputs the last 5 lines of the prior output to the terminal. By adjusting the arguments we can change the file, the number of lines starting from the beginning output and the number of lines from the end of that, which are output to the terminal.
  "$1","$2", and "$3" refer to the arguments taken by the shell script. "$1" is the first argument (in this case pride_and_prejudice.txt). "$2" and "$3" refer to the second and third arguments (20 and 5 respectively from the example). This allows the script to take a variety of arguments.

### Q4. More fun with Linux

Try these commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

**Solution:**

cal displays the calendar for the current month
``` {bash}
cal
```

cal 2021 displays the calendar for 2021  
``` {bash}
cal 2021
```
cal 9 1752 displays the calendar for September 1752. September 3rd through September 13th are missing from this month is 1752.
``` {bash}
cal 9 1752
```
date lists the current date and time 
``` {bash}
date
```
hostname shows the system's host name
``` {bash}
hostname
```
arch shows the cpu architecture of the host computer
``` {bash}
arch
```
uname -a prints all system information from the uname command, including the Kernel information, processor information, and machine name
``` {bash}
uname -a
```
uptime lists how long since the last system reboot
``` {bash}
uptime
```
whoami prints the userid of the user
``` {bash}
whoami
```
who prints info about the currently logged in users (who they are, when and where they logged in from)
``` {bash}
who
```
w prints information about the currently logged in users, what they are doing and uptime information
``` {bash}
w
```
id prints the user id and group ids for the user
``` {bash}
id
```
last | head shows a list of the last ten logins to the computer
``` {bash}
last | head
```
echo {con,pre}{sent,fer}{s,ed} outputs a list of all possible combinations of the listed strings in each set of brackets, remaining in the same order.
``` {bash}
echo {con,pre}{sent,fer}{s,ed}
```
time sleep 5 shows how long it takes to execute the sleep 5 command, which tells the system to wait for 5 seconds before continuing
``` {bash}
time sleep 5
```
history | tail lists the last ten commands inputted into the terminal
``` {bash}
set -o history
echo 'hi'
history | tail
```

